Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-07-12 manuscript/manuscript.tex
Continue writing Theory

\documentclass[extra]{gji}
%\documentclass[extra, referee]{gji}

\usepackage[utf8]{inputenc}
\usepackage{timet}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{todonotes} % to make annotations on margins

@@

\title[
    Variable Density Tesseroids]
    {Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth
}
\author[S.R. Soler, L. Uieda and M.E. Gimenez]
    {Santiado R. Soler$^{1,2}$, Leonardo Uieda$^3$ and Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina. e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de San Juan, Argentina\\
    $^3$Universidade do Estado do Rio de Janeiro, Rio de Janeiro, Brazil
    }

@@

Lithosphere density variation with depth has been studied for almost a century. 
\citet{Athy1930} obtained a decreasing exponential relation between both quantities by studying shale samples.
In the following years, other density functions have been proposed for different rock types \citep[e.g.][]{Maxant1980, Rao1986, Rao1993, Rao1994}.
Furthermore, the density variation with depth has been taken into account in forward and inversion gravity models, mostly applied to grabens and basins \citep{Cordell1973, Rao1986, Cowie1990, Rao1993, Rao1994, Zhang2001, Welford2010}.

@@

\begin{equation}
    V(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \frac{\rho(r')}{\ell} \kappa \,  dr' d\phi' d\lambda',
\label{eq:tesspot}
\end{equation}
\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') \frac{\Delta_\alpha}{\ell^3}
    \kappa \, dr' d\phi' d\lambda',
\label{eq:tessgrav}
\end{equation}
\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') I_{\alpha\beta} \, \kappa \, dr' d\phi' d\lambda' ,
    \label{eq:tesstensor}
\end{equation}
\noindent where
\begin{equation}
    I_{\alpha\beta} =
    \left(
        \frac{3\Delta_{\alpha} \Delta_{\beta}}{\ell^5} -
        \frac{\delta_{\alpha\beta}}{\ell^3}
    \right) ,
    \label{eq:tesstensorkernel}
\end{equation}

@@

According to \citet[p.390]{Hildebrand1987}, we can approximate any integral in the interval $[-1, 1]$ using a $N$th order GLQ, i.e. by a weighted sum of the integration kernel evaluated on the roots of the $N$th Legendre polynomial.

@@

\noindent where $x_i$ are the roots of the $N$th Legendre polynomial $P_N(x)$ and $W_i$ are the weights of each term.

@@

In case our integration is defined on a different interval, e.g. $[a,b]$, we can scale the nodes and perform a similar approximation.

@@

We are going to use only a second order GLQ, so the roots of the $P_2(x)$ Legendre polynomial are $x_i = \pm 0.577350269$ and the weights are $W_i = 1$.

Our intention is to use GLQ to approximate the gravity fields volume integrals from equations \ref{eq:tesspot}-\ref{eq:tesstensor}. Applying the GLQ definition for every integral corresponding to each spherical coordinate we obtain the following expression \citep{Asgharzadeh2007, Uieda2016}:

\begin{equation}
    \iiint\limits_\Omega f(r', \phi', \lambda') d\Omega \approx A 
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda f(r_i, \phi_j, \lambda_k),
\end{equation}

@@

In our case with variable density, we can write the integral kernel as:

\begin{equation}
    f(r', \phi', \lambda') = \rho(r') g(r', \phi', \lambda')
\end{equation}

>>>>

\documentclass[extra]{gji}
%~ %\documentclass[extra, referee]{gji}

\usepackage[utf8]{inputenc}
\usepackage{timet}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{todonotes} % to make annotations on margins

@@

\title[Variable Density Tesseroids]{
    Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth
}
\author[S.R. Soler, L. Uieda and M.E. Gimenez]{
    Santiado R. Soler$^{1,2}$, Leonardo Uieda$^3$ and Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina. e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de San Juan, Argentina\\
    $^3$Universidade do Estado do Rio de Janeiro, Rio de Janeiro, Brazil
    }

@@

Lithosphere density variation with depth has been studied for almost a century. 
\citet{Athy1930} obtained a decreasing exponential relation between both quantities by studying shale samples.
In the following years, other density functions have been proposed for different rock types \citep[e.g.,][]{Maxant1980, Rao1986, Rao1993, Rao1994}.
Furthermore, the density variation with depth has been taken into account in forward and inversion gravity models, mostly applied to grabens and basins \citep{Cordell1973, Rao1986, Cowie1990, Rao1993, Rao1994, Zhang2001, Welford2010}.

@@

\begin{equation}
    V(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \frac{\rho(r')}{\ell} \kappa \,  dr' d\phi' d\lambda',
\label{eq:tesseroid-pot}
\end{equation}
\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') \frac{\Delta_\alpha}{\ell^3}
    \kappa \, dr' d\phi' d\lambda',
\label{eq:tesseroid-grav}
\end{equation}
\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') I_{\alpha\beta} \, \kappa \, dr' d\phi' d\lambda' ,
    \label{eq:tesseroid-tensor}
\end{equation}
\noindent where
\begin{equation}
    I_{\alpha\beta} =
    \left(
        \frac{3\Delta_{\alpha} \Delta_{\beta}}{\ell^5} -
        \frac{\delta_{\alpha\beta}}{\ell^3}
    \right) ,
    \label{eq:tesseroid-tensor-kernel}
\end{equation}

@@

According to \citet[p.~390]{Hildebrand1987}, we can approximate any integral in the interval $[-1, 1]$ using a $N$th order GLQ, i.e. by a weighted sum of the integration kernel evaluated on the roots of the $N$th Legendre polynomial:

@@

\noindent where $x_i$ are the roots of the $N$th Legendre polynomial $P_N(x)$ and $W_i$ are the weights of each term:

@@

In case of an integration defined on a arbitrary interval, e.g. $[a,b]$, we can scale the nodes and perform a similar approximation:

@@

For example, if we use a second order GLQ, the roots of the $P_2(x)$ Legendre polynomial and its corresponding weights are $x_i = \pm 0.577350269$ and $W_i = 1$, respectivetly.

Our intention is to use GLQ to approximate the volume integrals from equations \ref{eq:tesseroid-pot}-\ref{eq:tesseroid-tensor}. 
When tesseroids have variable densities, we can write the integral kernels as the product of a certain function $g$ and the density function:

\begin{equation}
    f(r', \phi', \lambda') = \rho(r') g(r', \phi', \lambda'),
\end{equation}

\noindent and then apply the quadrature to every integral corresponding to each spherical coordinate, obtaining a similar expression to the one shown by \citet{Asgharzadeh2007} and \citet{Uieda2016}:

\iftwocol{
\begin{equation}
\begin{split}
    \iiint\limits_\Omega \rho(r') g(r', \phi', \lambda') d\Omega \approx& \\
    A 
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    & W_i^r W_j^\phi W_k^\lambda \rho(r_i) g(r_i, \phi_j, \lambda_k),
\end{split}
\label{eq:glq-var-dens}
\end{equation}
}
{
\begin{equation}
    \iiint\limits_\Omega \rho(r') g(r', \phi', \lambda') d\Omega \approx
    A 
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda \rho(r_i) g(r_i, \phi_j, \lambda_k),
\label{eq:glq-var-dens}
\end{equation}
}

@@

From equation \ref{eq:glq-var-dens} we can observe that the GLQ approximates the gravity fields of a tesseroid with variable density as the ones generated by point masses located at the nodes of the Legendre polynomials.
Furthermore, the information of the density function is summarised as the values it assumes in those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with a well fitted discretization algorithm, approximates the gravity fields with good accuracy and precision.\todo{no se si poner esta ultima oracion}
